---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
First, load the required libraries that are needed later.
```{r load_packages, message=FALSE, warning=FALSE, results='hide'}
chooseCRANmirror(graphics=FALSE, ind=1)
if (!("dplyr" %in% library())) install.packages("dplyr",type="binary",quiet=TRUE)
library(dplyr)
if (!("lattice" %in% library())) install.packages("lattice",quiet=TRUE)
library(lattice)
```

## Loading and preprocessing the data
Load the data:
```{r loading_data}
con <- unz(description="activity.zip",filename="activity.csv")
activity <- read.csv(con)
```

Explore the data:
```{r str}
str(activity)
```

```{r summary}
summary(activity)
```
Further preprocessing of the data is not necessary.

## What is mean total number of steps taken per day?
Calculate the total number of steps taken each day. `NA`'s are removed.
```{r create_steps_per_day_data}
steps_per_day <- activity %>%
    group_by(date) %>%
    summarise(steps=sum(steps,na.rm=TRUE))
```

Plot the resulting data.
```{r barplot_steps_per_day}
barplot(steps~date, steps_per_day, main="Total number of steps taken each day")
```

Calculate the mean and median of the total number of steps for all days.
```{r calc_mean_median_of_tot_steps_per_day}
mean_total_daily_steps <- mean(steps_per_day$steps)
mean_total_daily_steps
median_total_daily_steps <- median(steps_per_day$steps)
median_total_daily_steps
```
The **mean** of the total daily steps is `r format(mean_total_daily_steps,nsmall=0)` and the **median** of the total daily steps is
`r median_total_daily_steps`.

## What is the average daily activity pattern?
Calculate the number of steps in each 5-minute interval averaged over all days in the dataset. `Na`'s are removed.
```{r create_daily_mean_interval_data}
steps_by_interval <- activity %>%
    group_by(interval) %>%
    summarise(steps=mean(steps,na.rm=TRUE))
```

Plot the resulting data.
```{r lineplot_steps_by_interval}
plot(steps~interval, steps_by_interval, type="l", main="Daily average number of steps in each 5-minute interval")
```

Calculate the 5-minute interval that, on average over all days, contains the maximum number of steps.
```{r calc_int_of_max_average_steps}
max_int <- which.max(steps_by_interval$steps)
max_interval <- steps_by_interval[max_int,]$interval
max_interval
```
The 5-minute interval that on average over all days contains the maximum number of steps is `r max_interval`.

## Imputing missing values
Calculation of the total number of missing values in the dataset.
```{r calc_NAs}
total_NA <- colSums(is.na(activity))
```
There are `r total_NA[1]` rows with `NA`'s in the `activity` dataset.

I decided to replace missing values by the mean steps of the corresponding interval.
```{r replace_NAs}
activity_noNAs <- activity
for (i in 1:nrow(activity_noNAs)) {
    if (is.na(activity_noNAs[i,"steps"])) {
        impute_steps <- steps_by_interval[steps_by_interval$interval==activity_noNAs[i,"interval"],"steps"]
        activity_noNAs[i,"steps"] <- as.integer(impute_steps)
    }
}
```
Calculate the total steps per day on the dataset with imputed NAs.
```{r create_steps_per_day_noNAs}
steps_per_day_noNAs <- activity_noNAs %>%
    group_by(date) %>%
    summarise(steps=sum(steps))
```
The corresponding plot is as follows:
```{r barplot_noNAs}
barplot(steps~date, steps_per_day_noNAs, main="Total number of steps taken each day")
```

Calculate the mean and median of the total number of steps for all days.
```{r calc_mean_median_with_noNAs}
mean_total_daily_steps_noNAs <- mean(steps_per_day_noNAs$steps)
mean_total_daily_steps_noNAs
median_total_daily_steps_noNAs <- median(steps_per_day_noNAs$steps)
median_total_daily_steps_noNAs
```
The **mean** of the total daily steps is `r format(mean_total_daily_steps_noNAs,nsmall=0)` and the **median** of the total daily steps is
`r median_total_daily_steps_noNAs`. These values differ from the case with NAs ignored.

```{r comparison_table}
t <- data.frame(mean  =c(  mean_total_daily_steps,  mean_total_daily_steps_noNAs),
                median=c(median_total_daily_steps,median_total_daily_steps_noNAs)
                ,row.names=c("NAs ignored","NAs imputed")
                )
t
```
Apparently the mean and median with NAs imputed are bigger than those with NAs ignored.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable `daytype` in the dataset `activity_noNAs` with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r create_daytype_variable}
Sys.setlocale("LC_ALL","English_US")
activity_noNAs$daytype <- factor(ifelse(weekdays(as.Date(activity_noNAs$date)) %in% c("Saturday","Sunday"), "weekend", "weekday") )
str(activity_noNAs$daytype)
```
Calculate the number of steps in each 5-minute interval averaged over weekday and weekend.
```{r create_daily_mean_interval_data_noNAs}
steps_by_interval_noNAs <- activity_noNAs %>%
    group_by(interval,daytype) %>%
    summarise(steps=mean(steps))
```
The time series of the daily-averaged 5-minute interval steps of weekdays and weekends is shown below:
```{r lattice_daytype}
xyplot(steps ~ interval | daytype, steps_by_interval_noNAs, type="l", layout=c(1,2))
```

